Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-07 source/paper.tex
almost finished abstract

\begin{abstract}
TEXT
\end{abstract}

@@

\subsection{Viscous and diffusive terms}
\label{sec: diffusion}
For the simulations presented here we do not require viscous or
diffusive terms for stability, but sometimes they are required to
produce converged solutions so that refining the grid does not lead to
smaller and smaller scale motions. Viscous or diffusive terms involve
the laplacian of the prognostic field, which in all cases is not
globally defined for the finite element function spaces we use in this
paper. Here we present the symmetric interior penalty discretisation
method \citep{cockburn2007note} applied to the viscous term that we
add to the prognostic velocity equation when required. This method can
be adapted to apply to scalar fields when required.

>>>>

\begin{abstract}
Gusto is a python code library for building dynamical cores that are
the heart of weather and climate prediction models. Gusto is built on
top of Firedrake which provides the software infrastructure required
to solve partial differential equations using the finite element
method. The compressible Euler equations solved by the dynamical core
have a particular mathematical structure that should be preserved by
the spatial discretisation. Recent research has shown that one way to
do this is using compatible finite element methods which also have the
advantage that this structure is preserved even on non-orthogonal
grids, enabling better parallel scaling through spatial domain
decomposition. This was the motivation for the development of the Met
Office new dynamical core, LFRic. However, there are different
families of compatible finite element method that can be defined on
different grids, such as the cubed sphere or icosahedral sphere
grids. For each family the different prognostic fields are discretised
using finite element spaces that have different orders and
interelement continuity. These function spaces can be difficult to
construct, in particular on the sphere where the Piola transform is
required to ensure that the horizontal wind vector is tangential to
the sphere. In addition, when simulating flows that have a much larger
horizontal scale than vertical scale, it is advantageous to have
meshes that are aligned in vertical columns; Firedrake provides such
meshes and Gusto provides the correct finite element families for the
compatible finite element method and a range of algorithms for stable
and accurate timestepping fields discretised on these finite element
spaces.

Gusto is intended as a tool for rapid prototyping of new algorithms,
based on compatible finite element methods, for dynamical cores. In
this context, the development of new algorithms typically follows a
path where they are tested on simpler equation sets such as the
shallow water equations or the Boussinesq equations. These equation
sets share important features of the mathematical structure of the
compressible Euler equation set and so provide a good indication of
whether new algorithms will work in a dynamical core, but they provide
a simpler setting for investigating the properties of the
algorithms. Therefore Gusto provides compatible finite element spatial
discretisations of these equation sets and then the timestepping
algorithms are written in such a way that they can be applied to any
of these equation sets. ** write a bit more about what this enables?
** talk about physics?

In this paper we describe the algorithms implemented in Gusto and
demonstrate their accuracy (and efficiency?) when applied to the
hierarchy of test problems commonly used in the numerical weather
prediction community.
\end{abstract}

@@

\subsection{Viscous and diffusive terms}
\label{sec: diffusion}
For the simulations presented here we do not require viscous or
diffusive terms for stability, but sometimes they are required to
produce converged solutions so that refining the grid does not lead to
smaller and smaller scale motions. Viscous or diffusive terms involve
the laplacian of the prognostic field, which for all fields is not
globally defined for the finite element function spaces we use in this
paper. In Gusto we use the symmetric interior penalty discretisation
method \citep{cockburn2007note} which we descirbe below applied to the
viscous term added to the prognostic velocity equation when
required. This method can be adapted to apply to scalar fields.
